My title

### 2024-08-08 paper/paper.qmd
Rearrange data section

# Data {#sec-data}
## Sources
A good indicator of inflation is the Consumer Price Index (CPI), as it measures the change in price using a fixed basket of goods and services. Statistics Canada is a government agency that collects information on Canada’s economy (cite). Thus, it has the data that provides the CPI for numerous years. Statistics Canada also provides us with other datasets such as the average household income and the average household expenditures, in a year for all the provinces. We have chosen to look at Ontario solely, as we hold firmly Ontario behaves very differently due to its high population in relation to the other provinces. The data for Canada as a whole is complicated as there are many missing values for some provinces. Our analysis and the data provided by Statistics Canada makes it possible to reproduce our model for different provinces using the github repository.  While there are other datasets available, these three datasets are provided by the same source and has data for provinces across Canada. The expenditure and cpi data also provides the information for three additional specific categories, food, shelter, and recreation, making it easier to isolate and answer our research question. @sec-variables will illustrate the visualised data.

@@

@@

## Variables {#sec-variables}

    x = "Year",
    y = "Expenditure Value ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = spending_colours, name = NULL) +
  scale_x_continuous(
    breaks = seq(2010, 2021, by = 2)) +
  scale_y_continuous(
      labels = scales::comma
    )
```
From the household spending dataset, we chose to select the values for all food expenditures, all shelter expenditures, recreation expenditures, and total expenditures. This choice was made because these are principal expenditures for everyone. Recreation was added to this mix, to identify how consumers behave when inflation is high. @fig-food_n_recreation indicates this relationship. In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth. Note for simplicity, we did not include the other categories used for total expenditures. For all four of the categories, the mean and median values were close to each other, indicating the data is not skewed and did not have many outliers. However, the standard deviation for shelter and total expenditures were higher which indicates more money was spent and thus, more variation over the years. The only variable we use for our models from this dataset is `total expenditure`. However, the model can be used for the other categories. The model is made to be easily reproducible and can be found in the github link provided on page 1.

@@

From the CPI dataset, we chose to look at the annual average to keep it consistent with the annual average of expenditures. The data provides cpi for specific categories, as well as all-items. We chose to select cpi for all-items, food, shelter, and recreations, for comparison with the household spending. The prices for food and shelter increased almost 35% in the 10 years illustrated in @fig-cpi. However, the prices for recreation saw a decrease in 2020 and 2021, which indicates a relation to COVID-19. Our models will use the ‘all-items’ CPI to ensure consistency.

@@

From the income dataset, we chose to isolate the values for the category “all persons with income”. This indicate all tax fillers and dependants that earn income from all sources. This was done to keep our analysis consistent, as we could not separate who was spending. Our model will use the all persons with income variable as well. @fig-income shows a significant increase in the number of income earners. This could be due to a population increase, specifically an increase in tax fillers and tax dependants.

>>>>

# Data {#sec-data}

## Sources and Variables {#sec-sources_variables}
A good indicator of inflation is the Consumer Price Index (CPI), as it measures the change in price using a fixed basket of goods and services. Statistics Canada is a government agency that collects information on Canada’s economy (cite). Thus, it has the data that provides the CPI for numerous years. Statistics Canada also provides us with other datasets such as the average household income and the average household expenditures, in a year for all the provinces. We have chosen to look at Ontario solely, as we hold firmly Ontario behaves very differently due to its high population in relation to the other provinces. The data for Canada as a whole is complicated as there are many missing values for some provinces. Our analysis and the data provided by Statistics Canada makes it possible to reproduce our model for different provinces using the github repository.  While there are other datasets available, these three datasets are provided by the same source and has data for provinces across Canada. The expenditure and cpi data also provides the information for three additional specific categories, food, shelter, and recreation, making it easier to isolate and answer our research question. @sec-sources_variables will illustrate the visualised data.

@@

From this dataset, we chose to select the values for all food expenditures, all shelter expenditures, recreation expenditures, and total expenditures. This choice was made because these are principal expenditures for everyone. Recreation was added to this mix, to identify how consumers behave when inflation is high. @fig-food_n_recreation indicates this relationship. In @fig-spending, we observe a steady increase in recreation expenditures, while food, shelter, and overall expenditures show a more rapid growth. Note for simplicity, we did not include the other categories used for total expenditures. For all four of the categories, the mean and median values were close to each other, indicating the data is not skewed and did not have many outliers. However, the standard deviation for shelter and total expenditures were higher which indicates more money was spent and thus, more variation over the years. The only variable we use for our models from this dataset is `total expenditure`. However, the model can be used for the other categories. The model is made to be easily reproducible and can be found in the github link provided on page 1.

    x = "Year",
    y = "Expenditure Value ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = spending_colours, name = NULL) +
  scale_x_continuous(
    breaks = seq(2010, 2021, by = 2)) +
  scale_y_continuous(
      labels = scales::comma
    )
```

@@

From this dataset, we chose to look at the annual average to keep it consistent with the annual average of expenditures. The data provides cpi for specific categories, as well as all-items. We chose to select cpi for all-items, food, shelter, and recreations, for comparison with the household spending. The prices for food and shelter increased almost 35% in the 10 years illustrated in @fig-cpi. However, the prices for recreation saw a decrease in 2020 and 2021, which indicates a relation to COVID-19. Our models will use the ‘all-items’ CPI to ensure consistency.

@@

From this dataset, we chose to isolate the values for the category “all persons with income”. This indicate all tax fillers and dependants that earn income from all sources. This was done to keep our analysis consistent, as we could not separate who was spending. Our model will use the all persons with income variable as well. @fig-income shows a significant increase in the number of income earners. This could be due to a population increase, specifically an increase in tax fillers and tax dependants.

@@
